PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2022-09-08 supplementary.tex
Update on Overleaf.

We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC). The number of isolates selected per species, average chromosome length and average pairwise divergence on core genes are reported in Table~\ref{table:panx-dataset}. These genomes had previously been analyzed with PanX \cite{ding2018panx}.

% TODO: also report average core genome size?
% TODO: evaluate core genome divergence? Or average mash triangle distance?
% TODO: avoid talking about PanX?
% TODO: add link with accesion numbers and download?

\begin{table}[hb]
    \begin{tabular}{l l l l}
        \hline\hline
        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
        \hline
        E.~Coli         & 307         & 5.0                          &                                       \\
        K.~Pneumoniae   & 109         & 5.3                          &                                       \\
        H.~Pylori       & 85          & 1.6                          &                                       \\
        M.~Tuberculosis & 51          & 4.4                          &                                       \\
        P.~Marinus      & 10          & 1.8                          &                                       \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset overview}. We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species. For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
    \label{table:panx-dataset}
\end{table}

@@

In Fig.~\ref{fig:panx-benchmark-suppl} we report the results of the benchmark. Results are divided on every panel by species, with different colors indicating the five possible choices for the alignment kernel.\\
Execution times (cf. panel B) vary from few minutes for PM to several hours for EC. This is consistent with dataset sizes. Aligning sequences with \textit{mmseqs2} consistently requires more time than \textit{minimap2}. The time required to build a pangraph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}. The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is not particularly restrictive when aligning several tens of isolates.\\
% TODO: describe the properties of the graph (panel D-E-F)
The total pangenome length (panel G) depends both on the number of isolates and the ability of the alignment kernel to merge homologous sequences. When setting the energy parameters to zero more diverged sequences are merged into the same pancontigs and the total pangenome graph size is consistently smaller. The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression of 1-2\% for graphs with a high number of isolates (EC) or with small accessory genome (MT). The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction. For HP the highest core fraction is reached when using null energy parameters combined with \textit{mmseqs2} as alignment kernel, as this allows the merging of more diverged homologous sequences. For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence.

@@

% TODO: explain the sense of L50 and N50

@@

% TODO: simplify dataset descritpion (is already described above)

@@

\bibliography{cite}{}

>>>>

We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC). In Table~\ref{table:panx-dataset} we report some summary statistics for these datasets. The number of isolates per species ranges from several hundred isolates for EC to 10 for PM.
These data were previously analyzed using PanX \cite{ding2018panx}, a computational pipeline to cluster homologous genes together. This allowed us to identify core genes, and evaluate their average pairwise sequence divergence $\avg{d}$. Moreover we evaluate the fraction $f_p$ of core genes in the pangenome (weighted by gene length) and the average fraction $f_g$ of core gene sequence in a single genome.\\
EC is the dataset with the highest number of isolates (307) and is also the one with the biggest accessory genome ($f_p \sim 4\%$). MT is the species with the smallest sequence divergence ($\avg{d} \sim 0.03\%$) and the biggest core genome ($f_p \sim 58\%$). The two datasets with highest sequence divergence are HP ($\avg{d} \sim 4\%$) and PM ($\avg{d} \sim 24\%$). For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel. The sequence divergence of PM is instead too high and cannot be handled correctly by any of the kernels.

\begin{table}[hb]
    \setlength{\tabcolsep}{9pt}
    \begin{tabular}{l l c c c c}
        \hline\hline
        Species         & N. isolates & $\avg{l}$ [Mbp] & $\avg{d}$ & $f_p$  & $f_g$  \\
        \hline
        E.~Coli         & 307         & 5.0             & 1.6\%     & 3.7\%  & 15.4\% \\
        K.~Pneumoniae   & 109         & 5.3             & 0.8\%     & 17.4\% & 46.5\% \\
        H.~Pylori       & 85          & 1.6             & 4.2\%     & 30.2\% & 42.3\% \\
        M.~Tuberculosis & 51          & 4.4             & 0.03\%    & 57.9\% & 61.7\% \\
        P.~Marinus      & 10          & 1.8             & 24.3\%    & 21.0\% & 45.6\% \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset overview}. We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species. For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{l}$, average pairwise divergence on core genes $\avg{d}$, fraction of core gene sequence in the pangenome $f_p$ and fraction of core gene sequence in a single genome $f_g$.}
    \label{table:panx-dataset}
\end{table}

@@

In Fig.~\ref{fig:panx-benchmark-suppl} we report the results of the benchmark. Results are divided on every panel by species, with different colors indicating the five possible choices for the alignment kernel.\\
Execution times (cf. panel B) vary from few minutes for PM to several hours for EC. This is consistent with dataset sizes. Aligning sequences with \textit{mmseqs2} consistently requires more time than \textit{minimap2}. The time required to build a pangraph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}. The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is not particularly restrictive when aligning several tens of isolates.\\
We quantify the properties of pancontings by looking at their total number (panel D) the minimum number of pancontigs that include at least 50\% of all the pangenome graph sequence (L50 statistics, panel E) and the minimum length of pancontigs in this minimal set (N50 statistics, panel F). As expected, the use of null energy parameter results in general in a more fragmented graph, with a higher number of pancontigs that on average have a shorter length. HP is an exception, since it includes some highly-diverged sequences and the use of null energy parameters increases the default threshold divergence $d \sim 10\%$. This allows merging of sequences with higher divergence by \textit{mmseqs2} kernel, but not by the default \textit{minimap2} kernel, whose threshold average divergence is itself around 10\% (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}). A second exception is PM. The average divergence of this dataset is far above the capabilities of the alignment kernels, and only the use of \textit{mmseqs2} with null energy parameters allows for a partial merging of some homologous sequences. The resulting graph is composed of few very long pancontigs, mostly containing sequence from a single isolate. If we exclude this exception, in general the pangenome graphs of our datasets have around $10^3$ / $10^4$ pancontigs, with 50\% of the sequence being contained in around 10\% of them, having minimum length of 1-10 kbp.\\
The total pangenome length (panel G) depends both on the number of isolates and the ability of the alignment kernel to merge homologous sequences. When setting the energy parameters to zero more diverged sequences are merged into the same pancontigs and the total pangenome graph size is consistently smaller. The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression of 1-2\% for graphs with a high number of isolates (EC) or with small accessory genome (MT). The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction. For HP the highest core fraction is reached when using null energy parameters combined with \textit{mmseqs2} as alignment kernel, as this allows the merging of more diverged homologous sequences. For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence, but is still unable to merge highly diverged sequences.

@@

% TODO: explain the meaning of L50 and N50

@@

@@

\bibliography{cite}
